A Template for Two or One-Column Vignettes

### 2020-08-12 manuscript/manuscript.Rmd
pdf version

---
title: "Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures"
author:
  - Aaron Conway PhD:
      email: aaron.conway@utoronto.ca
      phone: +1 416 946 7112
      address: 585 University Ave | Toronto, ON  M5G 2N2
      institute: [PMCC, LBFON, QUT]
      correspondence: true
  - Kristina Chang MScN:
      institute: [PMCC]
      correspondence: false
  - Megan Bittner BSc:
      institute: [LBFON]
      correspondence: false
  - Dan Phan BSc:
      institute: [LBFON]
      correspondence: false
  - Navpreet Kamboj BScN:
      institute: [LBFON]
      correspondence: false
  - Matteo Parotto MD, PhD:
      institute: [TGH, UoT]
      correspondence: false
  - Amanda Matthews:
      institute: [PMCC]
      correspondence: false
  - Sherryl Alexander:
      institute: [PMCC]
      correspondence: false
  - Pauline Glaves:
      institute: [PMCC]
      correspondence: false
  - Shahvand Masihi:
      institute: [IR]
      correspondence: false
  - Sebastian Mafeld:
      institute: [IR]
      correspondence: false
institute:
  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
  - UoT: Department of Anesthesia and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
  - IR: Interventional Radiology, JDMI, Toronto General Hospital
bibliography: Extras/references.bib
csl: Extras/sp-basic.csl
output:
  officedown::rdocx_document:
    reference_docx: Extras/manuscript_template.docx
    pandoc_args:
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

library(drake)
library(magrittr)
library(flextable)

@@

\pagebreak

@@

A prospective observational study was conducted. Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included. Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale. Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity items.

@@

Despite the relevance and value of assessing thirst-discomfort of patients, the subjective experience of thirst presents challenges in developing a valid and reliable tool to succinctly and accurately measure its symptoms and level of discomfort prior to procedures. A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The seven-item perioperative thirst-discomfort scale (PTDS-7) was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients. Six items on the scale had a weighted kappa coefficient of 1, while the item "I feel like drinking water" was 0.97.[@Martins_2017] Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91.[@Martins_2017]

Although the initial validation of the PTDS-7 was conducted in a different context (i.e. *post*-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use. The aim of this study was to examine the validity and reliability of the PTDS-7 for measuring thirst discomfort for patients who are fasting *before* medical procedures.

@@

Mokken scaling is a method for establishing if items in a scale conform to a cumulative, hierarchical structure.[@perng2012construct] An intentionally simple example of a hierarchical scale is a set of items enquiring about height that increases in increments of 10cm from 100cm to 200cm. In this example, taller people would endorse more items than those who were shorter.[@watson2008hierarchy]

To determine if items conform to a hierarchical structure, Mokken scaling analyses properties of individual items as described by the item characteristic curve. Item characteristic curves show the association between the score on an item to the level of the latent trait being measured. Mokken scaling makes only the assumption of monotone homogeneity about the association between the score on an item and the level of the latent trait. This means that as the trait increases, so does the item score. "Difficulty", is a term used in Mokken Scale Analysis to refer to the extent to which items are endorsed by respondents. Items at the upper end of the range of the latent trait are characterized as being more difficult. Items in a Mokken Scale are arranged along the latent trait in terms of their difficulty. The properties of items can be measured using the scalability coefficient, *H* (Loevinger's coefficient). *H* measures the extent to which items are arranged as expected by their mean values along the latent trait. *H* > 0.3 is the minimum acceptable value, with values > 0.4 indicating a moderate scale and > 0.5 indicates a strong scale.

@@

Mokken scale analysis was conducted in this study to explore whether there were hierarchical properties in participant ratings of thirst discomfort and to explore the dimensions of the PTDS. Mokken scale analysis proceeded by first checking the PTDS-7 scalability coefficients. If **H~i~** was below 0.3 or if the lower limit of the 95% CI (confidence interval) for *H~i~* was below 0.3, the item was to be excluded. Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified. To assess for violations of invariant item ordering, crit values <40 were considered acceptable.[@molenaar2000mps5] Provided invariant item ordering holds for a set of items, the coefficient **H~T~** expresses the accuracy of the ordering, with values below 0.3 being unacceptable and above 0.4 indicating moderate invariant item ordering.[@Ligtvoet_2010] The coefficient score (rho), which is similar to Cronbach's alpha, was used to assess reliability of the scale.

@@

No items required removal from the PTDS-7 based on the condition that scalability coefficients (*H~i~* and lower bound of 95% CI) should be over 0.3 (Appendix Table \@ref(tab:Hvaluesptds7)). We used a lower bound for *c*, starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore the dimensionality of the PTDS-7. From 0.05 to 0.45, all of the items formed a single scale. The item, *'I have a bad taste in my mouth'*, was excluded by selecting the remaining items which showed uni-dimensionality at a threshold level of .50. The 6 remaining items (PTDS-6) were further examined for the criteria of a Mokken scale and item invariant ordering. The PTDS-6 item set has a homogeneity value *H(se)* of `r readd(H_values_ptds6)$H`, which indicated a strong Mokken scale. Results indicated that there were no significant violations of the assumption of monotonicity (z-sig). As such, no items were removed based on this criteria.

@@

@@

\pagebreak

## References

<div id="refs"></div>

\pagebreak

## Figure legend

Fig. \@ref(fig:distribution) Distribution of PTDS-5 scores

Fig. \@ref(fig:intensity) Association between PTDS-5 and thirst intensity

Fig. \@ref(fig:globalthirst) Association between PTDS-5 and global thirst rating

Fig. \@ref(fig:fasting) Association between PTDS-5 and duration of fasting

\pagebreak

```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
drake::loadd(summary_table)
gtsummary::as_flextable(summary_table)
```

\pagebreak

```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
  autofit()
```

@@

```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
knitr::include_graphics("figures/distribution.png")
```

\pagebreak

```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
knitr::include_graphics("figures/thirst-intensity.png")
```

\pagebreak

```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
knitr::include_graphics("figures/global-thirst.png")
```

\pagebreak

```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
knitr::include_graphics("figures/fasting.png")
```

\pagebreak

## Appendix

```{r tab.id='Hvaluesptds7', tab.cap.pre="Appendix Table ", tab.cap = "Homogeneity values (and standard errors) for all items in the PTDS"}
drake::readd(H_values_ptds7_table) %>% 
  flextable::flextable() %>% 
  autofit()
```

\pagebreak

```{r tab.id='monotonicity', tab.cap.pre="Appendix Table ", tab.cap="monotonicity with default minsize"}

flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
  autofit()
```

\pagebreak

```{r tab.id='aisp', tab.cap.pre="Appendix Table ", tab.cap = "aisp for increasing H thresholds (*c*)"}
flextable::flextable(drake::readd(aisp)) %>% 
  autofit()
```

\pagebreak

```{r tab.id='iioptds6', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-6"}
drake::readd(iio_ptds6_table) %>% 
  flextable() %>% 
  autofit()
```

\pagebreak

```{r tab.id='iioptds5', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-5"}
drake::readd(iio_ptds5_table) %>%
    flextable() %>%
    autofit()
```

<!---BLOCK_LANDSCAPE_STOP--->

>>>>

---
title: A Template for Two or One-Column Vignettes

# Use letters for affiliations
author:
  - name: ""
#     affiliation: a
#   - name: Second Author
#     affiliation: a,b
# address:
#   - code: a
#     address: Institute of Smoke and Magic, University of Sometown, Sometown, XY, 12345
#   - code: b
#     address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890

# Optional: line of arbitrary text with additional information.
# Could be used, for example, to mention the bibliographic info in a post-print.
# If not specified, defaults to "This version was compiled on \today"
#date_subtitle: Published in *Journal of Statistical Software*, 2018

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Author and Author

# Place eg a DOI URL or CRAN Package URL here
doi_footer: "https://cran.r-project.org/package=YourPackage"

# Abstract
abstract: |
  Your abstract will be typeset here, and used by default a visually distinctive font.
  An abstract should explain to the general reader the major contributions of the article.

# Optional: Acknowledgements
acknowledgements: |
  This template package builds upon, and extends, the work of the excellent
  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
  gratefully acknowledged as this work would not have been possible without them.  Our extensions
  are under the same respective licensing term
  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).

# Optional: One or more keywords
keywords:
  - one
  - two
  - optional
  - keywords
  - here

# Paper size for the document, values of letter and a4
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
#one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
#numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography 
bibliography: references

# Optional: Enable a 'Draft' watermark on the document
#watermark: true

# Customize footer, eg by referencing the vignette
footer_contents: "Perioperative thirst discomfort scale"

# Produce a pinp document
output: pinp::pinp

---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.retina = 5)
library(drake)
library(magrittr)
# library(flextable)

@@

@@

A prospective observational study was conducted. Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included. Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale. Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity ratings.

@@

Despite the relevance and value of assessing thirst-discomfort of patients, the subjective experience of thirst presents challenges in developing a valid and reliable tool to succinctly and accurately measure its symptoms and level of discomfort prior to procedures. A thirst-discomfort scale for perioperative use was developed at a surgical centre affiliated with an accredited public hospital in Brazil.[@Martins_2017] The seven-item perioperative thirst-discomfort scale (PTDS) was developed in three stages, including face and content validation and reliability, and based on the Consensus-based Standards for the selection of health status Measurement Instruments (COSMIN) checklist.[@mokkink2010cosmin] Inter-rater reliability was tested through inter-observer equivalence where a pair of nurses independently, but simultaneously administered the scale among 70 patients. Six items on the scale had a weighted kappa coefficient of 1, while the item "I feel like drinking water" was 0.97.[@Martins_2017] Cronbach's alpha was used to test the internal consistency of the scale with a value of 0.91.[@Martins_2017]

Although the initial validation of the PTDS was conducted in a different context (i.e. *post*-surgical patients), it demonstrates strong potential for accurately assessing thirst-discomfort more generally for peri-operative use. The aim of this study was to examine the validity and reliability of the PTDS for measuring thirst discomfort for patients who are fasting *before* medical procedures.

@@

Mokken scaling is a method for establishing if items in a scale conform to a cumulative, hierarchical structure.[@perng2012construct] When an instrument contains hierarchically ordered items, the items can be ordered from those indicating mild to severe symptoms. An intentionally simple example of a hierarchical scale is a set of items enquiring about height that increases in increments of 10cm from 100cm to 200cm. In this example, taller people would endorse more items than those who were shorter.[@watson2008hierarchy]

To determine if items conform to a hierarchical structure, Mokken scaling analyses properties of individual items as described by the item characteristic curve. Item characteristic curves show the association between the score on an item to the level of the latent trait being measured. Mokken scaling makes only the assumption of monotone homogeneity about the association between the score on an item and the level of the latent trait. This means that as the trait increases, so does the item score. "Difficulty", is a term used in Mokken Scale Analysis to refer to the extent to which items are endorsed by respondents. Items at the upper end of the range of the latent trait are characterized as being more difficult. Items in a Mokken Scale are arranged along the latent trait in terms of their difficulty. The properties of items can be measured using the scalability coefficient, *H* (Loevinger's coefficient). *H* measures the extent to which items are arranged as expected by their mean values along the latent trait. *H* \> 0.3 is the minimum acceptable value, with values \> 0.4 indicating a moderate scale and \> 0.5 indicates a strong scale.

@@

Mokken scale analysis was conducted in this study to explore whether there were hierarchical properties in participant ratings of thirst discomfort and to explore the dimensions of the PTDS. Mokken scale analysis proceeded by first checking the PTDS-7 scalability coefficients. If **H~i~** was below 0.3 or if the lower limit of the 95% CI (confidence interval) for *H~i~* was below 0.3, the item was to be excluded. Then, scale partitioning was carried out to explore the dimensions of the PTDS through increasing c (lower bound c defines the minimum value of coefficients **H~i~** in the Mokken scale by 0.05 increments).[@molenaar2000mps5] Monotone homogeneity and invariant item ordering were investigated at the whole scale level, as no sub-scales were identified. To assess for violations of invariant item ordering, crit values \<40 were considered acceptable.[@molenaar2000mps5] Provided invariant item ordering holds for a set of items, the coefficient **H~T~** expresses the accuracy of the ordering, with values below 0.3 being unacceptable and above 0.4 indicating moderate invariant item ordering.[@Ligtvoet_2010] The coefficient score (rho), which is similar to Cronbach's alpha, was used to assess reliability of the scale.

@@

No items required removal from the PTDS-7 based on the condition that scalability coefficients (*H~i~* and lower bound of 95% CI) should be over 0.3 (Appendix Table \@ref(tab:Hvaluesptds7)). We used a lower bound for *c*, starting from 0.05 and increasing to 0.80 in 0.05 increments, to explore the dimensionality of the PTDS-7. From 0.05 to 0.45, all of the items formed a single scale. The item, '*I have a bad taste in my mouth*', was excluded by selecting the remaining items which showed uni-dimensionality at a threshold level of .50. The 6 remaining items (PTDS-6) were further examined for the criteria of a Mokken scale and item invariant ordering. The PTDS-6 item set has a homogeneity value *H(se)* of `r readd(H_values_ptds6)$H`, which indicated a strong Mokken scale. Results indicated that there were no significant violations of the assumption of monotonicity (z-sig). As such, no items were removed based on this criteria.

@@

```{r }
drake::readd(item_hierarchy_df_ptds5) %>% 
  dplyr::select(item) 
```

@@

<!-- ## References -->

<!-- ::: {#refs} -->
<!-- ::: -->

<!-- ## Figure legend -->

<!-- Fig. \@ref(fig:distribution) Distribution of PTDS-5 scores -->

<!-- Fig. \@ref(fig:intensity) Association between PTDS-5 and thirst intensity -->

<!-- Fig. \@ref(fig:globalthirst) Association between PTDS-5 and global thirst rating -->

<!-- Fig. \@ref(fig:fasting) Association between PTDS-5 and duration of fasting -->

```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
drake::loadd(summary_table)
# gtsummary::as_flextable(summary_table)
gtsummary::as_kable_extra(summary_table, booktabs = T)
```

@@

```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
# knitr::include_graphics("figures/distribution.png")
drake::readd(distribution_plot)
```

```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
# knitr::include_graphics("figures/thirst-intensity.png")
drake::readd(thirst_intensity_plot)
```

```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
# knitr::include_graphics("figures/global-thirst.png")
library(dplyr)
drake::readd(global_thirst_plot)
```

```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
# knitr::include_graphics("figures/fasting.png")
drake::readd(fasting_duration_ptds_plot)
```
